---
title: "Integrative R Learner on NSW Data"
format:
  html:
    theme: cosmo
    embed-resources: true
    number-sections: true
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
    df-print: kable
---

```{r}
set.seed(123)
library(grf)
library(ggplot2)
```


# Load Data
```{r}
nsw_experimental <- read.csv('data/nsw_experimental.csv')
```

```{r}
nsw_observational <- read.csv('data/nsw_observational_resampling_trimmed_new.csv')
# nsw_observational <- read.csv('data/nsw_observational.csv')
```

```{r}
str(nsw_observational)
```


```{r}
str(nsw_experimental)
```

# Combine Data

Row bind the data and create an indicator variable (1 = experimental, 0 = observational)
```{r}
nsw_experimental$S <- 1
nsw_observational$S <- 0

data_all <- rbind(nsw_experimental, nsw_observational)
```


# Define Outcome, Covariates, Assignment, and Origin

```{r}
Y <- data_all$re78
A <- data_all$treat

X <- data_all[, c("age","educ","black","hisp","marr","nodegree","re74","re75")]
S <- data_all$S
```


# Calculate Propensity Score

```{r}
prop_model <- glm(
  S ~ .,
  data = data_all[, c("S", "age","educ","black","hisp","marr","nodegree","re74","re75")],
  family = binomial()
)

e_hat <- predict(
  prop_model,
  newdata = data_all[, c("age","educ","black","hisp","marr","nodegree","re74","re75")],
  type = "response"
)
```





## Check Overlap

```{r}
df <- data.frame(
  ps = e_hat,
  treat = data_all$treat
)

ggplot(df, aes(x = ps, fill = factor(treat))) +
  geom_density(alpha = 0.4) +
  labs(fill = "Treatment", x = "Propensity score", y = "Density")
```

Wow, look at that overlap. Even with the truth mixed in, we still have this mass on the left for control units




# Calculate Outcome Model

```{r}
# RCT outcome model
mu_rct_model <- lm(re78 ~ age + educ + black + hisp + marr + nodegree + re74 + re75,
                   data = nsw_experimental)

# Observational outcome model
mu_obs_model <- lm(re78 ~ age + educ + black + hisp + marr + nodegree + re74 + re75,
                   data = nsw_observational)

X_rct <- X[S == 1, ]
X_obs <- X[S == 0, ]

mu_hat <- numeric(nrow(X))

mu_hat[S == 1] <- predict(mu_rct_model, newdata = X_rct)
mu_hat[S == 0] <- predict(mu_obs_model, newdata = X_obs)
```





# Initial Confounding Function

This could be any value, but to have a smart initialization, we use the difference in outcomes between observed and RCT

```{r}
mu_rct_pred <- predict(mu_rct_model, newdata = X)
mu_obs_pred <- predict(mu_obs_model, newdata = X)

c_hat <- mu_obs_pred - mu_rct_pred
```


# Fit Tau using Causal Forest

```{r}
eps_Y <- Y - mu_hat
eps_A <- A - e_hat
weights <- eps_A^2
```


```{r}
for (b in 1:4) {

  Y_tilde <- (eps_Y / eps_A) - (1 - S) * c_hat

  tau_forest <- regression_forest(
    X,
    Y_tilde,
    sample.weights = weights
  )

  tau_hat <- predict(tau_forest, X)$predictions


  Y_tilde <- (eps_Y / eps_A) - tau_hat

  obs_idx <- which(S == 0)

  c_forest <- regression_forest(
    X[obs_idx,],
    Y_tilde[obs_idx],
    sample.weights = weights[obs_idx]
  )

  c_hat <- predict(c_forest, X)$predictions
}
```


```{r}
tau_hat <- predict(tau_forest, X)$predictions
```


Get ATE
```{r}
ate <- mean(tau_hat)  # or all X, but no weights
ate
```

```{r}
ate_se <- sd(tau_hat) / sqrt(length(tau_hat))
ate_se
```

