---
title: "R Learner and Integrative R Learner on NSW Data"
format:
  html:
    theme: cosmo
    embed-resources: true
    number-sections: true
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
    df-print: kable
---

This file contains our experiments on R-Learner on both observational and experimental data as well as an Integrative R-Learner on the combined dataset. If you haven't already, please run Required.R to ensure packages run smoothly. 

Please also change the file paths for the dataset here:

```{r}
experimental_path <- 'data/nsw_experimental.csv'
observational_path <- 'data/nsw_observational_resampling_trimmed_new.csv'
```



```{r}
set.seed(123)
library(knitr)
library(grf)
library(ggplot2)
library(caret)
```

k-fold indicator
```{r}
use_k_fold <- TRUE
```


# Load Data
```{r}
nsw_experimental <- read.csv(experimental_path)
```

```{r}
nsw_observational <- read.csv(observational_path)
```

```{r}
str(nsw_observational)
```


```{r}
str(nsw_experimental)
```


# R-Learner Implementation


## Transform Data

```{r}
features <- c("age","educ","black","hisp","marr","nodegree","re74","re75")

Y_exp <- nsw_experimental$re78
A_exp <- nsw_experimental$treat

X_exp <- nsw_experimental[, features]

###
Y_obs <- nsw_observational$re78
A_obs <- nsw_observational$treat

X_obs <- nsw_observational[, features]

n_exp <- nrow(nsw_experimental)
n_obs <- nrow(nsw_observational)
K <- 5
```

```{r}
# k fold indicator
use_k_fold <- TRUE
```


## Calculate Propensity Scores


```{r}
# propensity score
if (use_k_fold == FALSE){
  prop_model_exp <- glm(
    A ~ .,
    data = data.frame(A = A_exp, X_exp),
    family = binomial()
  )
  prop_model_obs <- glm(
    A ~ .,
    data = data.frame(A = A_obs, X_obs),
    family = binomial()
  )
  
  e_hat_exp <- predict(
    prop_model_exp,
    newdata = X,
    type = "response"
  )
  
  e_hat_obs <- predict(
    prop_model_obs,
    newdata = X,
    type = "response"
  )
}
```

```{r}
if (use_k_fold == TRUE){
  n_exp <- nrow(nsw_experimental)
  n_obs <- nrow(nsw_observational)
  K <- 5
  
  fold_id_exp <- sample(rep(1:K, length.out = n_exp))
  fold_id_obs <- sample(rep(1:K, length.out = n_obs))
  
  e_hat_exp <- rep(NA, n_exp)
  e_hat_obs <- rep(NA, n_obs)

  features <- c("age","educ","black","hisp","marr","nodegree","re74","re75")
  
  for (k in 1:K) {
  
    train_idx_exp <- which(fold_id_exp != k)
    test_idx_exp  <- which(fold_id_exp == k)
    
    train_idx_obs <- which(fold_id_obs != k)
    test_idx_obs  <- which(fold_id_obs == k)
  
    train_data_exp <- data.frame(A = A_exp[train_idx_exp], X_exp[train_idx_exp, ])
    test_data_exp  <- X_exp[test_idx_exp, ]
    
    train_data_obs <- data.frame(A = A_obs[train_idx_obs], X_obs[train_idx_obs, ])
    test_data_obs  <- X_obs[test_idx_obs, ]
  
    # fit model on training fold
    prop_model_k_exp <- glm(
      A ~ .,
      data = train_data_exp,
      family = binomial()
    )
    prop_model_k_obs <- glm(
      A ~ .,
      data = train_data_obs,
      family = binomial()
    )
  
    # predict on held-out fold
    e_hat_exp[test_idx_exp] <- predict(
      prop_model_k_exp,
      newdata = test_data_exp,
      type = "response"
    )
    e_hat_obs[test_idx_obs] <- predict(
      prop_model_k_obs,
      newdata = test_data_obs,
      type = "response"
    )
  }
}
```


## Calculate Outcome Models


```{r}
# outcome model
if (use_k_fold == FALSE) {
  mu_model_exp <- lm(re78 ~ ., data = data.frame(re78 = Y_exp, X_exp))
  mu_hat_exp <- predict(mu_model_exp, newdata = X_exp)
  
  mu_model_obs <- lm(re78 ~ ., data = data.frame(re78 = Y_obs, X_obs))
  mu_hat_obs <- predict(mu_model_obs, newdata = X_obs)
}
```


```{r}
if (use_k_fold == TRUE){
  n_exp <- nrow(nsw_experimental)
  n_obs <- nrow(nsw_observational)
  K <- 5
  
  mu_hat_exp <- rep(NA, n_exp)
  mu_hat_obs <- rep(NA, n_obs)
  
  # fold_id <- sample(rep(1:K, length.out = n))
  
  for (k in 1:K) {
  
    train_idx_exp <- which(fold_id_exp != k)
    test_idx_exp  <- which(fold_id_exp == k)
    
    train_idx_obs <- which(fold_id_obs != k)
    test_idx_obs  <- which(fold_id_obs == k)
  
    fit_exp <- lm(re78 ~ ., data = data.frame(re78 = Y_exp[train_idx_exp], 
                                              X_exp[train_idx_exp, ]))
    fit_obs <- lm(re78 ~ ., data = data.frame(re78 = Y_obs[train_idx_obs], 
                                              X_obs[train_idx_obs, ]))
  
    mu_hat_exp[test_idx_exp] <- predict(fit_exp, newdata = X_exp[test_idx_exp, ])
    mu_hat_obs[test_idx_obs] <- predict(fit_obs, newdata = X_obs[test_idx_obs, ])
  }
}
```


## Estimate Tau's using Weighted regressions

```{r}
# used Claude to help generate code
# r loss minimization
# residuals
Y_tilde_exp <- Y_exp - mu_hat_exp
A_tilde_exp <- A_exp - e_hat_exp

Y_tilde_obs <- Y_obs - mu_hat_obs
A_tilde_obs <- A_obs - e_hat_obs

# if any A_tilde is close to 0, the weight W will blow up
A_tilde_exp <- ifelse(abs(A_tilde_exp) < 1e-6, 1e-6, A_tilde_exp)
A_tilde_obs <- ifelse(abs(A_tilde_obs) < 1e-6, 1e-6, A_tilde_obs)

# pseudo-outcome and weights
W_exp <- Y_tilde_exp / A_tilde_exp
weights_exp <- A_tilde_exp^2

W_obs <- Y_tilde_obs / A_tilde_obs
weights_obs <- A_tilde_obs^2

# weighted regression to estimate tau(X)
tau_model_exp <- lm(W ~ ., data = data.frame(W = W_exp, X_exp), weights = weights_exp)
tau_hat_exp <- predict(tau_model_exp, newdata = X_exp)

tau_model_obs <- lm(W ~ ., data = data.frame(W = W_obs, X_obs), weights = weights_obs)
tau_hat_obs <- predict(tau_model_obs, newdata = X_obs)

# summary
summary(tau_hat_exp)
mean(tau_hat_exp)  # average treatment effect estimate for experimental data

summary(tau_hat_obs)
mean(tau_hat_obs) # average treatment effect estimate for observational data
```
Results

```{r}
summary(tau_model_exp)
summary(tau_model_obs)
# Mean(\hat{\tau}) = -131.78 is the average treatment effect across individuals in our dataset. The difference in income between the treatment and control groups is $-131.78, so the training program decreased earnings, on average.
```

```{r}
hist(tau_hat_exp, xlab = "tau hat", main = "Estimated Treatment Effect Across Individuals for Experimental Data", breaks = 50)
hist(tau_hat_obs, xlab = "tau hat", main = "Estimated Treatment Effect Across Individuals for Observational Data", breaks = 50)
# For many individuals, it seems the treatment (training program) did not have much of an effect on their income, as the tau hat value for many individuals falls around 0. 
```

```{r}
se_exp <- sd(tau_hat_exp) / sqrt(length(tau_hat_exp))
se_exp
se_obs <- sd(tau_hat_obs) / sqrt(length(tau_hat_obs))
se_obs
```


# Integrative R-Learner

## Reload data to reset


```{r}
set.seed(123)
```

```{r}
nsw_experimental <- read.csv(experimental_path)
```

```{r}
nsw_observational <- read.csv(observational_path)
```

## Combine Data

Row bind the data and create an indicator variable (1 = experimental, 0 = observational)
```{r}
nsw_experimental$S <- 1
nsw_observational$S <- 0

data_all <- rbind(nsw_experimental, nsw_observational)
```


## Define Outcome, Covariates, Assignment, and Origin

```{r}
Y <- data_all$re78
A <- data_all$treat

X <- data_all[, c("age","educ","black","hisp","marr","nodegree","re74","re75")]
S <- data_all$S
```


## Calculate Propensity Score


```{r}
if (use_k_fold == FALSE){
  prop_model <- glm(
    S ~ .,
    data = data_all[, c("S", "age","educ","black","hisp","marr","nodegree","re74","re75")],
    family = binomial()
  )
  
  e_hat <- predict(
    prop_model,
    newdata = data_all[, c("age","educ","black","hisp","marr","nodegree","re74","re75")],
    type = "response"
  )
}
```


```{r}
if (use_k_fold == TRUE){
  n <- nrow(data_all)
  K <- 5
  
  fold_id <- sample(rep(1:K, length.out = n))
  
  e_hat <- rep(NA, n)

  features <- c("age","educ","black","hisp","marr","nodegree","re74","re75")
  
  for (k in 1:K) {
  
    train_idx <- which(fold_id != k)
    test_idx  <- which(fold_id == k)
  
    train_data <- data_all[train_idx, c("S", features)]
    test_data  <- data_all[test_idx, features]
  
    # fit model on training fold
    prop_model_k <- glm(
      S ~ .,
      data = train_data,
      family = binomial()
    )
  
    # predict on held-out fold
    e_hat[test_idx] <- predict(
      prop_model_k,
      newdata = test_data,
      type = "response"
    )
  }
}
```


### Check Overlap

```{r}
df <- data.frame(
  ps = e_hat,
  treat = data_all$treat
)

ggplot(df, aes(x = ps, fill = factor(treat))) +
  geom_density(alpha = 0.4) +
  labs(fill = "Treatment", x = "Propensity score", y = "Density")
```

Wow, look at that overlap. Even with the truth mixed in, we still have this mass on the left for control units




## Calculate Outcome Model


```{r}


if (use_k_fold == FALSE) {
  # RCT outcome model
  mu_rct_model <- lm(re78 ~ age + educ + black + hisp + marr + nodegree + re74 + re75,
                     data = nsw_experimental)
  
  # Observational outcome model
  mu_obs_model <- lm(re78 ~ age + educ + black + hisp + marr + nodegree + re74 + re75,
                     data = nsw_observational)
  
  X_rct <- X[S == 1, ]
  X_obs <- X[S == 0, ]
  
  mu_hat <- numeric(nrow(X))
  
  mu_hat[S == 1] <- predict(mu_rct_model, newdata = X_rct)
  mu_hat[S == 0] <- predict(mu_obs_model, newdata = X_obs)
}

```

```{r}
if (use_k_fold == TRUE){
  K <- 5
  
  mu_hat <- numeric(nrow(X))
  
  # -------------------------
  # RCT folds
  # -------------------------
  rct_idx <- which(S == 1)
  
  folds_rct <- createFolds(rct_idx, k = K, list = TRUE)
  
  for (k in 1:K) {
  
    test_idx <- rct_idx[folds_rct[[k]]]
    train_idx <- setdiff(rct_idx, test_idx)
  
    fit <- lm(
      re78 ~ age + educ + black + hisp + marr + nodegree + re74 + re75,
      data = data.frame(
        re78 = Y[train_idx],
        X[train_idx, ]
      )
    )
  
    mu_hat[test_idx] <- predict(
      fit,
      newdata = X[test_idx, ]
    )
  }
  
  # -------------------------
  # Observational folds
  # -------------------------
  obs_idx <- which(S == 0)
  
  folds_obs <- createFolds(obs_idx, k = K, list = TRUE)
  
  for (k in 1:K) {
  
    test_idx <- obs_idx[folds_obs[[k]]]
    train_idx <- setdiff(obs_idx, test_idx)
  
    fit <- lm(
      re78 ~ age + educ + black + hisp + marr + nodegree + re74 + re75,
      data = data.frame(
        re78 = Y[train_idx],
        X[train_idx, ]
      )
    )
  
    mu_hat[test_idx] <- predict(
      fit,
      newdata = X[test_idx, ]
    )
  }
}
```




## Initial Confounding Function

This could be any value, but to have a smart initialization, we use the difference in outcomes between observed and RCT

```{r}

# RCT outcome model
mu_rct_model <- lm(re78 ~ age + educ + black + hisp + marr + nodegree + re74 + re75,
                   data = nsw_experimental)

# Observational outcome model
mu_obs_model <- lm(re78 ~ age + educ + black + hisp + marr + nodegree + re74 + re75,
                   data = nsw_observational)

mu_rct_pred <- predict(mu_rct_model, newdata = X)
mu_obs_pred <- predict(mu_obs_model, newdata = X)

c_hat <- mu_obs_pred - mu_rct_pred
```


## Fit Tau using Weighted Forests

This is taken directly from the appendix algorithm 2

```{r}
eps_Y <- Y - mu_hat
eps_A <- A - e_hat
weights <- eps_A^2
```

Weighted Forest Regression, weights give us a signal of how much an observation can give us a heterogenous effect
```{r}
for (b in 1:4) {

  Y_tilde <- (eps_Y / eps_A) - (1 - S) * c_hat

  tau_forest <- regression_forest(
    X,
    Y_tilde,
    sample.weights = weights
  )

  tau_hat <- predict(tau_forest, X)$predictions


  Y_tilde <- (eps_Y / eps_A) - tau_hat

  obs_idx <- which(S == 0)

  c_forest <- regression_forest(
    X[obs_idx,],
    Y_tilde[obs_idx],
    sample.weights = weights[obs_idx]
  )

  c_hat <- predict(c_forest, X)$predictions
}
```


```{r}
obs_idx <- which(S == 0)
tau_hat <- predict(tau_forest, X)$predictions
tau_hat_obs <- tau_hat[obs_idx]
```

We only want to see the ATE for the observational data
```{r}
hist(
  tau_hat_obs,
  xlab = "tau hat",
  main = "Estimated Treatment Effect Across Individuals for Observational Data",
  breaks = 300,
  xlim = c(-20000, 20000)
)
```

Get ATE
```{r}
ate <- mean(tau_hat_obs)  # or all X, but no weights
ate
```

```{r}
ate_se <- sd(tau_hat) / sqrt(length(tau_hat))
ate_se
```


# Final Results

```{r}
# Values you already have
ate_int  <- 862.2943
se_int   <- 146.1605

ate_rct  <- 1695.154
se_rct   <- 67.43434

ate_obs  <- -2243.241
se_obs   <- 118.0976

# Create table
results <- data.frame(
  Method = c("Integrative R-Learner", "R-Learner (RCT)", "R-Learner (Obs)"),
  ATE = c(ate_int, ate_rct, ate_obs),
  SE = c(se_int, se_rct, se_obs)
)

# 95% CI (Wald)
z <- 1.96
results$CI_lower <- results$ATE - z * results$SE
results$CI_upper <- results$ATE + z * results$SE

kable(results, digits = 3, caption = "ATE Estimates with 95% Confidence Intervals")
```

The integrative R-Learner ATE was estimated only using observational data, we get a very confident positive ATE using Integrative R-Learner. The R-Learner using only observational data yields a negative ATE which does not reflect the RCT ATE. This suggests that the Integrative R-Learner outperforms R-Learner, but we would need access to an experimental and observational dataset.

