---
title: "Integrative R Learner on NSW Data"
format:
  html:
    theme: cosmo
    embed-resources: true
    number-sections: true
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
    df-print: kable
---

```{r}
set.seed(123)
library(grf)
library(ggplot2)
library(caret)
```

k-fold indicator
```{r}
use_k_fold <- TRUE
```


# Load Data
```{r}
nsw_experimental <- read.csv('data/nsw_experimental.csv')
```

```{r}
nsw_observational <- read.csv('data/nsw_observational_resampling_trimmed_new.csv')
# nsw_observational <- read.csv('data/nsw_observational2.csv')
```

```{r}
str(nsw_observational)
```


```{r}
str(nsw_experimental)
```

# Combine Data

Row bind the data and create an indicator variable (1 = experimental, 0 = observational)
```{r}
nsw_experimental$S <- 1
nsw_observational$S <- 0

data_all <- rbind(nsw_experimental, nsw_observational)
```


# Define Outcome, Covariates, Assignment, and Origin

```{r}
Y <- data_all$re78
A <- data_all$treat

X <- data_all[, c("age","educ","black","hisp","marr","nodegree","re74","re75")]
S <- data_all$S
```


# Calculate Propensity Score



```{r}
if (use_k_fold == FALSE){
  prop_model <- glm(
    S ~ .,
    data = data_all[, c("S", "age","educ","black","hisp","marr","nodegree","re74","re75")],
    family = binomial()
  )
  
  e_hat <- predict(
    prop_model,
    newdata = data_all[, c("age","educ","black","hisp","marr","nodegree","re74","re75")],
    type = "response"
  )
}
```


```{r}
if (use_k_fold == TRUE){
  n <- nrow(data_all)
  K <- 5
  
  fold_id <- sample(rep(1:K, length.out = n))
  
  e_hat <- rep(NA, n)

  features <- c("age","educ","black","hisp","marr","nodegree","re74","re75")
  
  for (k in 1:K) {
  
    train_idx <- which(fold_id != k)
    test_idx  <- which(fold_id == k)
  
    train_data <- data_all[train_idx, c("S", features)]
    test_data  <- data_all[test_idx, features]
  
    # fit model on training fold
    prop_model_k <- glm(
      S ~ .,
      data = train_data,
      family = binomial()
    )
  
    # predict on held-out fold
    e_hat[test_idx] <- predict(
      prop_model_k,
      newdata = test_data,
      type = "response"
    )
  }
}
```


## Check Overlap

```{r}
df <- data.frame(
  ps = e_hat,
  treat = data_all$treat
)

ggplot(df, aes(x = ps, fill = factor(treat))) +
  geom_density(alpha = 0.4) +
  labs(fill = "Treatment", x = "Propensity score", y = "Density")
```

Wow, look at that overlap. Even with the truth mixed in, we still have this mass on the left for control units




# Calculate Outcome Model


```{r}


if (use_k_fold == FALSE) {
  # RCT outcome model
  mu_rct_model <- lm(re78 ~ age + educ + black + hisp + marr + nodegree + re74 + re75,
                     data = nsw_experimental)
  
  # Observational outcome model
  mu_obs_model <- lm(re78 ~ age + educ + black + hisp + marr + nodegree + re74 + re75,
                     data = nsw_observational)
  
  X_rct <- X[S == 1, ]
  X_obs <- X[S == 0, ]
  
  mu_hat <- numeric(nrow(X))
  
  mu_hat[S == 1] <- predict(mu_rct_model, newdata = X_rct)
  mu_hat[S == 0] <- predict(mu_obs_model, newdata = X_obs)
}

```

```{r}
if (use_k_fold == TRUE){
  K <- 5
  
  mu_hat <- numeric(nrow(X))
  
  # -------------------------
  # RCT folds
  # -------------------------
  rct_idx <- which(S == 1)
  
  folds_rct <- createFolds(rct_idx, k = K, list = TRUE)
  
  for (k in 1:K) {
  
    test_idx <- rct_idx[folds_rct[[k]]]
    train_idx <- setdiff(rct_idx, test_idx)
  
    fit <- lm(
      re78 ~ age + educ + black + hisp + marr + nodegree + re74 + re75,
      data = data.frame(
        re78 = Y[train_idx],
        X[train_idx, ]
      )
    )
  
    mu_hat[test_idx] <- predict(
      fit,
      newdata = X[test_idx, ]
    )
  }
  
  # -------------------------
  # Observational folds
  # -------------------------
  obs_idx <- which(S == 0)
  
  folds_obs <- createFolds(obs_idx, k = K, list = TRUE)
  
  for (k in 1:K) {
  
    test_idx <- obs_idx[folds_obs[[k]]]
    train_idx <- setdiff(obs_idx, test_idx)
  
    fit <- lm(
      re78 ~ age + educ + black + hisp + marr + nodegree + re74 + re75,
      data = data.frame(
        re78 = Y[train_idx],
        X[train_idx, ]
      )
    )
  
    mu_hat[test_idx] <- predict(
      fit,
      newdata = X[test_idx, ]
    )
  }
}
```




# Initial Confounding Function

This could be any value, but to have a smart initialization, we use the difference in outcomes between observed and RCT

```{r}

# RCT outcome model
mu_rct_model <- lm(re78 ~ age + educ + black + hisp + marr + nodegree + re74 + re75,
                   data = nsw_experimental)

# Observational outcome model
mu_obs_model <- lm(re78 ~ age + educ + black + hisp + marr + nodegree + re74 + re75,
                   data = nsw_observational)

mu_rct_pred <- predict(mu_rct_model, newdata = X)
mu_obs_pred <- predict(mu_obs_model, newdata = X)

c_hat <- mu_obs_pred - mu_rct_pred
```


# Fit Tau using Weighted Forests

This is taken directly from the appendix algorithm 2

```{r}
eps_Y <- Y - mu_hat
eps_A <- A - e_hat
weights <- eps_A^2
```


```{r}
for (b in 1:4) {

  Y_tilde <- (eps_Y / eps_A) - (1 - S) * c_hat

  tau_forest <- regression_forest(
    X,
    Y_tilde,
    sample.weights = weights
  )

  tau_hat <- predict(tau_forest, X)$predictions


  Y_tilde <- (eps_Y / eps_A) - tau_hat

  obs_idx <- which(S == 0)

  c_forest <- regression_forest(
    X[obs_idx,],
    Y_tilde[obs_idx],
    sample.weights = weights[obs_idx]
  )

  c_hat <- predict(c_forest, X)$predictions
}
```


```{r}
obs_idx <- which(S == 0)
tau_hat <- predict(tau_forest, X)$predictions
tau_hat_obs <- tau_hat[obs_idx]
```

We only want to see the ATE for the observational data
```{r}
hist(
  tau_hat_obs,
  xlab = "tau hat",
  main = "Estimated Treatment Effect Across Individuals for Observational Data",
  breaks = 300,
  xlim = c(-20000, 20000)
)
```

Get ATE
```{r}
ate <- mean(tau_hat_obs)  # or all X, but no weights
ate
```

```{r}
ate_se <- sd(tau_hat) / sqrt(length(tau_hat))
ate_se
```



