---
title: "Integrative R-Learner: NSW Application"
author: "Integrative R-Learner Analysis"
format:
  html:
    theme: cosmo
    embed-resources: true
    number-sections: true
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
    df-print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE, message = FALSE, fig.width = 8,
  fig.height = 5
)
library(pacman)
p_load(glmnet, grf, ggplot2, dplyr, tidyr, kableExtra)
```

# Background

The National Supported Work (NSW) Demonstration (1975--1979) was a
randomized job-training program for disadvantaged workers. LaLonde
(1986) showed that replacing experimental controls with
non-experimental CPS controls caused standard regression to produce
biased estimates. We apply the **Integrative R-Learner** (Wu & Yang,
2022) to combine the RCT arm with the CPS-augmented observational
sample, estimating heterogeneous treatment effects while accounting
for unmeasured confounding in the observational data.

# Setup

```{r load-method}
source("method_initial.r")
```

```{r load-data}
nsw_exp <- read.csv("data/nsw_experimental.csv")
nsw_obs <- read.csv("data/nsw_observational.csv")
```

# Data Overview

```{r dimensions}
cat("Experimental (RCT):", nrow(nsw_exp), "obs\n")
cat("Observational (CPS):", nrow(nsw_obs), "obs\n")
```

```{r summary-experimental}
summary(nsw_exp)
```

```{r summary-observational}
summary(nsw_obs)
```

## Treatment Balance

```{r treatment-balance}
cat("=== RCT treatment counts ===\n")
print(table(nsw_exp$treat))
cat("\n=== Observational treatment counts ===\n")
print(table(nsw_obs$treat))
```

## Covariate Distributions

```{r covariate-comparison, fig.height=8}
covars <- c(
  "age", "educ", "black", "hisp",
  "marr", "nodegree", "re74", "re75"
)

par(mfrow = c(2, 4), mar = c(4, 4, 2, 1))
for (v in covars) {
  boxplot(
    list(
      RCT = nsw_exp[[v]],
      CPS = nsw_obs[[v]]
    ),
    main = v, col = c("#4A90D9", "#E8634F"),
    border = "#333333", las = 1
  )
}
```

# Naive Estimates

```{r naive-rct}
ate_rct <- mean(nsw_exp$re78[nsw_exp$treat == 1]) -
  mean(nsw_exp$re78[nsw_exp$treat == 0])
cat("Experimental ATE:", round(ate_rct, 2), "\n")
```

```{r naive-obs}
ate_obs <- mean(nsw_obs$re78[nsw_obs$treat == 1]) -
  mean(nsw_obs$re78[nsw_obs$treat == 0])
cat("Naive observational ATE:", round(ate_obs, 2), "\n")
```

```{r naive-ols}
fit_ols <- lm(
  re78 ~ treat + age + educ + black + hisp +
    marr + nodegree + re74 + re75,
  data = nsw_obs
)
cat(
  "OLS-adjusted observational ATE:",
  round(coef(fit_ols)["treat"], 2), "\n"
)
```

# Integrative R-Learner

## Fit the Model

```{r fit-irlearner}
set.seed(42)

covariates <- c(
  "age", "educ", "black", "hisp",
  "marr", "nodegree", "re74", "re75"
)

result <- integrative_rlearner(
  data_rct = nsw_exp,
  data_obs = nsw_obs,
  outcome = "re78",
  treatment = "treat",
  covariates = covariates,
  degree = 2,
  K_nuisance = 5,
  K_loss = 5,
  lambda_grid = 10^seq(-3, 2, length.out = 12),
  ratio_grid = c(0.01, 0.1, 0.5, 1, 2, 5, 10, 50),
  verbose = TRUE
)
```

## Summary

```{r model-summary}
summary(result)
```

## Average Treatment Effect

```{r ate-comparison}
n_rct <- sum(result$S == 1)

ate_irlearner_all <- mean(result$tau_hat)
ate_irlearner_rct <- mean(result$tau_hat[1:n_rct])

comparison <- data.frame(
  Method = c(
    "RCT (difference-in-means)",
    "Naive observational",
    "OLS-adjusted observational",
    "Integrative R-Learner (all)",
    "Integrative R-Learner (RCT sample)"
  ),
  ATE = round(c(
    ate_rct, ate_obs, coef(fit_ols)["treat"],
    ate_irlearner_all, ate_irlearner_rct
  ), 2)
)
comparison
```

# Heterogeneous Treatment Effects

## CATE Distribution

```{r cate-distribution, fig.height=5}
tau_rct <- result$tau_hat[result$S == 1]
tau_obs <- result$tau_hat[result$S == 0]

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

hist(
  tau_rct, breaks = 30, col = "#4A90D980",
  border = "#4A90D9",
  main = "CATE in RCT sample",
  xlab = "Estimated CATE", las = 1
)
abline(v = mean(tau_rct), col = "#C0392B", lwd = 2)

hist(
  tau_obs, breaks = 30, col = "#E8634F80",
  border = "#E8634F",
  main = "CATE in OS sample",
  xlab = "Estimated CATE", las = 1
)
abline(v = mean(tau_obs), col = "#C0392B", lwd = 2)
```

## CATE by Subgroups

```{r cate-subgroups}
dat <- result$data
dat$tau_hat <- result$tau_hat
dat$c_hat <- result$c_hat

subgroup_cate <- function(var, label) {
  rct_sub <- dat[dat$S == 1, ]
  data.frame(
    Variable = label,
    Group_0 = round(
      mean(rct_sub$tau_hat[rct_sub[[var]] == 0]), 2
    ),
    Group_1 = round(
      mean(rct_sub$tau_hat[rct_sub[[var]] == 1]), 2
    ),
    Diff = round(
      mean(rct_sub$tau_hat[rct_sub[[var]] == 1]) -
        mean(rct_sub$tau_hat[rct_sub[[var]] == 0]), 2
    )
  )
}

rbind(
  subgroup_cate("black", "Black"),
  subgroup_cate("hisp", "Hispanic"),
  subgroup_cate("marr", "Married"),
  subgroup_cate("nodegree", "No Degree")
)
```

```{r cate-by-age, fig.height=5}
rct_dat <- dat[dat$S == 1, ]

par(mar = c(4, 4, 3, 1))
plot(
  rct_dat$age, rct_dat$tau_hat,
  pch = 16, col = "#4A90D980", cex = 0.8,
  xlab = "Age", ylab = "Estimated CATE",
  main = "CATE vs Age (RCT sample)",
  las = 1
)
lo <- loess(tau_hat ~ age, data = rct_dat)
age_seq <- seq(
  min(rct_dat$age), max(rct_dat$age),
  length.out = 100
)
lines(
  age_seq, predict(lo, data.frame(age = age_seq)),
  col = "#C0392B", lwd = 2
)
abline(h = 0, lty = 2, col = "gray50")
```

```{r cate-by-education, fig.height=5}
par(mar = c(4, 4, 3, 1))
plot(
  rct_dat$educ, rct_dat$tau_hat,
  pch = 16, col = "#27AE6080", cex = 0.8,
  xlab = "Education (years)",
  ylab = "Estimated CATE",
  main = "CATE vs Education (RCT sample)",
  las = 1
)
lo_ed <- loess(tau_hat ~ educ, data = rct_dat)
ed_seq <- seq(
  min(rct_dat$educ), max(rct_dat$educ),
  length.out = 100
)
lines(
  ed_seq, predict(lo_ed, data.frame(educ = ed_seq)),
  col = "#C0392B", lwd = 2
)
abline(h = 0, lty = 2, col = "gray50")
```

```{r cate-by-earnings, fig.height=5}
par(mar = c(4, 4, 3, 1))
plot(
  rct_dat$re75, rct_dat$tau_hat,
  pch = 16, col = "#8E44AD80", cex = 0.8,
  xlab = "Earnings in 1975 ($)",
  ylab = "Estimated CATE",
  main = "CATE vs Prior Earnings (RCT sample)",
  las = 1
)
lo_re <- loess(tau_hat ~ re75, data = rct_dat)
re_seq <- seq(
  min(rct_dat$re75), max(rct_dat$re75),
  length.out = 100
)
lines(
  re_seq, predict(lo_re, data.frame(re75 = re_seq)),
  col = "#C0392B", lwd = 2
)
abline(h = 0, lty = 2, col = "gray50")
```

# Confounding Function

```{r confounding-function}
obs_dat <- dat[dat$S == 0, ]
cat("Confounding function c(X) summary (OS only):\n")
print(summary(obs_dat$c_hat))
```

```{r confounding-hist, fig.height=4}
par(mar = c(4, 4, 3, 1))
hist(
  obs_dat$c_hat, breaks = 30,
  col = "#F39C1280", border = "#F39C12",
  main = "Confounding Function c(X) in OS",
  xlab = "Estimated c(X)", las = 1
)
abline(v = 0, lty = 2, col = "gray50", lwd = 2)
abline(
  v = mean(obs_dat$c_hat),
  col = "#C0392B", lwd = 2
)
```

# Comparison: RCT-Only vs Integrative

```{r rct-only-rlearner}
set.seed(42)

nuis_rct <- crossfit_nuisance(
  nsw_exp$re78, nsw_exp$treat,
  nsw_exp[, covariates], rep(1, nrow(nsw_exp)),
  K = 5
)
Y_res_rct <- nsw_exp$re78 - nuis_rct$mu_hat
A_res_rct <- nsw_exp$treat - nuis_rct$e_hat

B_rct <- build_poly_basis(
  nsw_exp[, covariates], degree = 2
)

best_cv_rct <- Inf
best_lambda_rct <- NA
for (lt in 10^seq(-3, 2, length.out = 12)) {
  n_rct_only <- nrow(nsw_exp)
  folds <- sample(rep(1:5, length.out = n_rct_only))
  cv_err <- numeric(5)
  for (k in 1:5) {
    tr <- folds != k
    te <- folds == k
    Z_tr <- A_res_rct[tr] * B_rct[tr, ]
    theta_k <- solve(
      crossprod(Z_tr) / sum(tr) +
        lt * diag(ncol(B_rct)),
      crossprod(Z_tr, Y_res_rct[tr]) / sum(tr)
    )
    fitted_te <- A_res_rct[te] *
      (B_rct[te, ] %*% theta_k)
    cv_err[k] <- mean(
      (Y_res_rct[te] - fitted_te)^2
    )
  }
  if (mean(cv_err) < best_cv_rct) {
    best_cv_rct <- mean(cv_err)
    best_lambda_rct <- lt
  }
}

Z_full <- A_res_rct * B_rct
alpha_rct <- solve(
  crossprod(Z_full) / nrow(nsw_exp) +
    best_lambda_rct * diag(ncol(B_rct)),
  crossprod(Z_full, Y_res_rct) / nrow(nsw_exp)
)

tau_rct_only <- as.numeric(B_rct %*% alpha_rct)
```

```{r compare-methods, fig.height=5}
par(mar = c(4, 4, 3, 1))
plot(
  tau_rct_only, tau_rct,
  pch = 16, col = "#2C3E5080", cex = 0.8,
  xlab = "RCT-only R-Learner CATE",
  ylab = "Integrative R-Learner CATE",
  main = "RCT-only vs Integrative R-Learner",
  las = 1
)
abline(0, 1, col = "#C0392B", lwd = 2, lty = 2)
abline(h = 0, v = 0, col = "gray70", lty = 3)
```

```{r variance-comparison}
var_rct_only <- var(tau_rct_only)
var_integrative <- var(tau_rct)

cat("Variance of RCT-only CATE:", round(var_rct_only, 2))
cat(
  "\nVariance of Integrative CATE:",
  round(var_integrative, 2)
)
cat(
  "\nEfficiency ratio (RCT/Integrative):",
  round(var_rct_only / var_integrative, 3), "\n"
)
```

# Summary

```{r final-summary}
cat("====================================\n")
cat("  Final Results Summary\n")
cat("====================================\n\n")
cat("Benchmark (RCT ATE):", round(ate_rct, 2), "\n")
cat(
  "Integrative R-Learner ATE:",
  round(ate_irlearner_rct, 2), "\n"
)
cat(
  "Naive observational ATE:",
  round(ate_obs, 2), "\n\n"
)
cat(
  "The Integrative R-Learner recovers an ATE",
  "close to the\nexperimental benchmark while",
  "the naive observational estimate\nis heavily",
  "biased by unmeasured confounding.\n"
)
```
